---
title: "MapReduce"
format: 
  pdf: 
    include-in-header: 
      text: '\pagenumbering{arabic}'
execute: 
  echo: false
  warning: false
---

```{r}
library(tidyverse)
library(here)
library(gt)
```

# Classroom example

This is the file to create a map reduce classroom example.
The setup is as follows

1. print this file
2. give the raw file to 1-4 students, they shall count the words. Measure the time. This is sequential processing.
3. give the chunked file to the respective number of students. There are $n = 2,4,5,10$ groups available. The number of students is the chunk server.
4. one student acts as the "reduce" server, collecting all the counts.
5. write the counts on the whiteboard including the times
6. which number of groups is the most efficient and which produces considerable overhead?

```{r}
clean <- function(x) {
  gsub('[,.;:\'"()]','',tolower(x))
}

word_count_simple <- function(lines) {
  chunks <- strsplit(clean(lines),'\\s')
  words <- do.call(c, chunks)
  table(words)
}

lorem <- read_lines(here("data","exchange.txt"))

clean_lorem <- clean(lorem)

chunks_lorem <- strsplit(clean_lorem,'\\s')

vec_lorem <- unlist(chunks_lorem)

two_groups <- split(vec_lorem, 1:2)

four_groups <- split(vec_lorem, 1:4)

five_groups <- split(vec_lorem, 1:5)

ten_groups <- split(vec_lorem, 1:10)

twenty_groups <- split(vec_lorem, 1:20)

```

## some summary statistics on the raw data

```{r}
lorem_freq_tbl <- word_count_simple(lorem)

lorem_df <- lorem_freq_tbl %>%
as.data.frame() %>%
mutate(
percentage = Freq/sum(Freq),
percentage_round = round(percentage,digits = 4)
) |>
arrange(desc(Freq))



```

In total there are `{r} sum(lorem_df$Freq)` words available.

```{r}
lorem_df |> gt() |> 
  fmt_percent(columns = c(percentage,percentage_round))
```

```{r}
#| fig-width: 10
#| fig-height: 15

lorem_df |> 
  ggplot(
    aes(
      y = reorder(words,-Freq),
      x = Freq
    )
  )+
  geom_col(
    color = "white"
  )+
  labs(
    title = "absolute frequency of all words in text",
    y = "",
    x = "count"
  )

```

```{r}
#| fig-width: 10
#| fig-height: 15

lorem_df |> 
  filter(percentage_round>0.01) |>  
  ggplot(
    aes(
      y = reorder(words,-percentage_round),
      x = percentage_round
    )
  )+
  geom_col(
    color = "white"
  )+
  labs(
    title = "relative frequency of words >1% in text",
    y = "",
    x = "percentage"
  )+
  scale_x_continuous(
    labels = scales::percent
  )

```

```{r}
#| fig-width: 10
#| fig-height: 15

lorem_df |> 
  filter(percentage_round>0.02) |>  
  ggplot(
    aes(
      y = reorder(words,-percentage_round),
      x = percentage_round
    )
  )+
  geom_col(
    color = "white"
  )+
  labs(
    title = "relative frequency of words >1% in text",
    y = "",
    x = "percentage"
  )+
  scale_x_continuous(
    labels = scales::percent
  )

```
{{< pagebreak >}}

# The Raw data 01

`{r} lorem`

{{< pagebreak >}}

# The Raw data 02

`{r} lorem`

{{< pagebreak >}}

# The Raw data 03

`{r} lorem`

{{< pagebreak >}}

# The Raw data 04

`{r} lorem`

{{< pagebreak >}}

# Two Groups

## Part 1 of 2 (`{r} length(two_groups[[1]])` words)

`{r} paste(two_groups[[1]],collapse = " ")`

{{< pagebreak >}}

## Part 2 of 2 (`{r} length(two_groups[[1]])` words)

`{r} paste(two_groups[[2]],collapse = " ")`

{{< pagebreak >}}

# Four Groups

## Part 1 of 4 (`{r} length(four_groups[[1]])` words)

`{r} paste(four_groups[[1]],collapse = " ")`

{{< pagebreak >}}

## Part 2 of 4 (`{r} length(four_groups[[1]])` words)

`{r} paste(four_groups[[2]],collapse = " ")`

{{< pagebreak >}}

## Part 3 of 4 (`{r} length(four_groups[[1]])` words)

`{r} paste(four_groups[[3]],collapse = " ")`

{{< pagebreak >}}

## Part 4 of 4 (`{r} length(four_groups[[1]])` words)

`{r} paste(four_groups[[4]],collapse = " ")`

{{< pagebreak >}}

# Five Groups

## Part 1 of 5 (`{r} length(five_groups[[1]])` words)

`{r} paste(five_groups[[1]],collapse = " ")`

{{< pagebreak >}}

## Part 2 of 5 (`{r} length(five_groups[[1]])` words)

`{r} paste(five_groups[[2]],collapse = " ")`

{{< pagebreak >}}

## Part 3 of 5 (`{r} length(five_groups[[1]])` words)

`{r} paste(five_groups[[3]],collapse = " ")`

{{< pagebreak >}}

## Part 4 of 5 (`{r} length(five_groups[[1]])` words)

`{r} paste(five_groups[[4]],collapse = " ")`

{{< pagebreak >}}

## Part 5 of 5 (`{r} length(five_groups[[1]])` words)

`{r} paste(five_groups[[5]],collapse = " ")`

{{< pagebreak >}}

# Ten Groups

## Part 1 of 10 (`{r} length(ten_groups[[1]])` words)

`{r} paste(ten_groups[[1]],collapse = " ")`

{{< pagebreak >}}

## Part 2 of 10 (`{r} length(ten_groups[[1]])` words)

`{r} paste(ten_groups[[2]],collapse = " ")`

{{< pagebreak >}}

## Part 3 of 10 (`{r} length(ten_groups[[1]])` words)

`{r} paste(ten_groups[[3]],collapse = " ")`

{{< pagebreak >}}

## Part 4 of 10 (`{r} length(ten_groups[[1]])` words)

`{r} paste(ten_groups[[4]],collapse = " ")`

{{< pagebreak >}}

## Part 5 of 10 (`{r} length(ten_groups[[1]])` words)

`{r} paste(ten_groups[[5]],collapse = " ")`

{{< pagebreak >}}

## Part 6 of 10 (`{r} length(ten_groups[[1]])` words)

`{r} paste(ten_groups[[6]],collapse = " ")`

{{< pagebreak >}}

## Part 7 of 10 (`{r} length(ten_groups[[1]])` words)

`{r} paste(ten_groups[[7]],collapse = " ")`

{{< pagebreak >}}

## Part 8 of 10 (`{r} length(ten_groups[[1]])` words)

`{r} paste(ten_groups[[8]],collapse = " ")`

{{< pagebreak >}}

## Part 9 of 10 (`{r} length(ten_groups[[1]])` words)

`{r} paste(ten_groups[[9]],collapse = " ")`

{{< pagebreak >}}

## Part 10 of 10 (`{r} length(ten_groups[[1]])` words)

`{r} paste(ten_groups[[10]],collapse = " ")`

{{< pagebreak >}}

# Twenty Groups

## Part 1 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[1]],collapse = " ")`

{{< pagebreak >}}

## Part 2 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[2]],collapse = " ")`

{{< pagebreak >}}

## Part 3 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[3]],collapse = " ")`

{{< pagebreak >}}

## Part 4 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[4]],collapse = " ")`

{{< pagebreak >}}

## Part 5 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[5]],collapse = " ")`

{{< pagebreak >}}

## Part 6 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[6]],collapse = " ")`

{{< pagebreak >}}

## Part 7 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[7]],collapse = " ")`

{{< pagebreak >}}

## Part 8 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[8]],collapse = " ")`

{{< pagebreak >}}

## Part 9 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[9]],collapse = " ")`

{{< pagebreak >}}

## Part 10 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[10]],collapse = " ")`

{{< pagebreak >}}

## Part 11 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[11]],collapse = " ")`

{{< pagebreak >}}

## Part 12 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[12]],collapse = " ")`

{{< pagebreak >}}

## Part 13 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[13]],collapse = " ")`

{{< pagebreak >}}

## Part 14 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[14]],collapse = " ")`

{{< pagebreak >}}

## Part 15 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[15]],collapse = " ")`

{{< pagebreak >}}

## Part 16 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[16]],collapse = " ")`

{{< pagebreak >}}

## Part 17 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[17]],collapse = " ")`

{{< pagebreak >}}

## Part 18 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[18]],collapse = " ")`

{{< pagebreak >}}

## Part 19 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[19]],collapse = " ")`

{{< pagebreak >}}

## Part 20 of 20 (`{r} length(twenty_groups[[1]])` words)

`{r} paste(twenty_groups[[20]],collapse = " ")`